---
title: "Assignment 09: Data Scraping"
author: "Elise Boos"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
#check wd
getwd()

#load packages
require(tidyverse)
require(rvest)
require(lubridate)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
#set webpage variable reading in as url
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')
webpage
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
#scrape data
water.system.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name
  
pswid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pswid

ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max.withdrawals.mgd <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4
#create dataframe from scaped data
withdrawals <- data.frame("Month" = month(c(1,5,9,2,6,10,3,7,11,4,8,12)),
                     "Year" = rep(2020,12),               
                     "Water.System" = water.system.name,
                     "PSWID" = pswid,
                     "Ownership" = ownership,
                     "Max.Withdrawals.mgd" = as.numeric(max.withdrawals.mgd))
withdrawals_2020 <- withdrawals %>%
  mutate(Date = my(paste0(Month,"-", Year)))

withdrawals_2020

#5
#plot max withdrawals by month
ggplot(withdrawals_2020,aes(x=Date,y=Max.Withdrawals.mgd)) + 
  geom_point() +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2020 Water usage data for",water.system.name),
       subtitle = ownership,
       y="Withdrawal (mgd)",
       x="Month")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.
#create scrape it function
scrape.it <- function(the_year, pswid_year){
  #Get the proper url
  the_url <- read_html(
    paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', pswid_year,'&year=',the_year))

  #Fetch the website
water.system.name <- the_url %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
pswid <- the_url %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
ownership <- the_url %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
max.withdrawals.mgd <- the_url %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

  #Construct a dataframe from the values
  the_df <- tibble("Month" = month(c(1,5,9,2,6,10,3,7,11,4,8,12)),
                     "Water.System" = water.system.name,
                     "PSWID" = pswid,
                     "Ownership" = ownership,
                     "Max.Withdrawals.mgd" = as.numeric(max.withdrawals.mgd)) %>% 
   mutate(Year = the_year) %>%
   mutate(Date = my(paste0(Month,"-", Year)))
  
  return(the_df)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
#use function for durham and 2015
durham_2015 <- scrape.it(2015, '03-32-010')
durham_2015

#plot max withdrawals by month
ggplot(durham_2015,aes(x=Date,y=Max.Withdrawals.mgd)) + 
  geom_point() +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Water usage data for Durham"),
       subtitle = ownership,
       y="Withdrawal (mgd)",
       x="Month")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Durham.Asheville.2015.data}
#8
#asheville for 2015
asheville_2015 <- scrape.it(2015, '01-11-010')
asheville_2015

#bind durham and asheville dataframes
a_d_2015 <- rbind(durham_2015, asheville_2015)

#plot coloring by water system
ggplot(a_d_2015,aes(x=Date,y=Max.Withdrawals.mgd)) + 
  geom_point() +
  geom_line(aes(color = Water.System)) +
  labs(title = paste("2015 Water usage data"),
       y="Withdrawal (mgd)",
       x="Month")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9
#use lapply to run for 2010-2019
years <- seq(2010,2019)
asheville_2010_2019 <- lapply(X = years, FUN = scrape.it, pswid_year = '01-11-010') %>% bind_rows()

#plot 
ggplot(asheville_2010_2019,aes(x=Date,y=Max.Withdrawals.mgd)) + 
  geom_point() +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2010-2019 Water usage data Asheville"),
       y="Withdrawal (mgd)",
       x="Date")
```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

>yes over time there is an increase in water usage in Asheville. 